Pre-registration Study 3 \n Who does not trust basic science?

### 2024-04-17 exp_3/cleaning.Rmd
preregistration study 3

---
title: "Data cleaning Study 2"
output: 
  html_document: 
    keep_md: yes
---

@@

## Import data
```{r import, message=FALSE}
d <- read_csv("./data/qualtrics.csv")
names(d)
```

@@

There are 6 failed attention checks.

@@

```{r, message=FALSE}
prolific_demographics <- read_csv("./data/prolific_demographics.csv")

@@

# clean and re-shape
d_wide <- d %>% 
  # add an easy to read participant identifier
  mutate(id = 1:nrow(.)) %>%
  # make some character variables numeric
  mutate(across(c(starts_with("BCTI"), starts_with("SICBS"), starts_with("wgm"), 
                  starts_with("pew")), ~extract_numbers(.)
  )
  ) %>% 
  mutate(across(starts_with("CMQ"), as.numeric)
  ) %>% 
  # add average conspiracy measures 
  rowwise() %>% 
  mutate(BCTI_avg = mean(c_across(starts_with("BCTI")
                                  ),
                         na.rm = TRUE
                         ),
         CMQ_avg = mean(c_across(starts_with("CMQ")
                                  ),
                         na.rm = TRUE
                         )
         ) %>% 
  ungroup() 
```

@@

```{r}
d_long <- d_wide %>% 
  # bring to long format
  pivot_longer(cols = electrons_know:water_accept, 
               names_to = "subject_question", values_to = "correct_accept") %>% 
  # separate conditions into CONVERGENCE_OUTCOME_STIMULUS
  separate_wider_delim(subject_question, "_", names = c("subject", "question")
                       ) %>%
  # necessary because each participant only saw ten out of 11 questions
  drop_na(correct_accept) %>% 
  pivot_wider(names_from = question, values_from = correct_accept) %>% 
# create better variable names
  rename(knowledge = know, 
         acceptance = accept) 
```

@@

# make data frame with average particpant data and wide format
d_wide <- left_join(d_wide %>% 
                 select(-c(ends_with("know"), ends_with("accept"))), 
               d_avg, 
               by = "id")
```

>>>>

---
title: "Data cleaning Study 3"
output: 
  html_document: 
    keep_md: yes
---

@@

## Import data
```{r import, message=FALSE}
d <- read_csv("data/test_study_3.csv")
names(d)
```

@@

There are % failed attention checks.

@@

```{r, message=FALSE}
prolific_demographics <- read_csv("data/prolific_demographics.csv")

@@

transform_text <- function(x) {
  ifelse(str_detect(x, "^I agree"), "Yes", "No")
}

# clean and re-shape
d_wide <- d %>% 
  # add an easy to read participant identifier
  mutate(id = 1:nrow(.)) %>%
  # make some character variables numeric
  mutate(across(c(starts_with("BCTI"), starts_with("SICBS"), starts_with("wgm"), 
                  starts_with("pew")), ~extract_numbers(.)
  )
  ) %>% 
  mutate(across(starts_with("CMQ"), as.numeric)
  ) %>% 
  # clean the text for correction 
  mutate_at(vars(ends_with("_confirm")), ~ transform_text(.))%>%
  # add average conspiracy measures 
  rowwise() %>% 
  mutate(BCTI_avg = mean(c_across(starts_with("BCTI")
                                  ),
                         na.rm = TRUE
                         ),
         CMQ_avg = mean(c_across(starts_with("CMQ")
                                  ),
                         na.rm = TRUE
                         )
         ) %>% 
  ungroup() 
```

@@

We use the initial acceptance question and the confirmation question (only participants who did not accept the consensus for the intial question answered the confirmation question) to make a final acceptance question. This final acceptance will take on the initial answer in case of initial acceptance, and the answer of the confirmation question in case of initial non-acceptance. The final question will simply be named `acceptance`.

```{r}
d_long <- d_wide %>% 
  # bring to long format
  pivot_longer(cols = c(ends_with("know"), ends_with("accept"), ends_with("confirm")), 
               names_to = "subject_question", values_to = "answer") %>% 
  # separate conditions into CONVERGENCE_OUTCOME_STIMULUS
  separate_wider_delim(subject_question, "_", names = c("subject", "question")
                       ) %>%
  pivot_wider(names_from = question, values_from = answer) %>% 
# create better variable names
  rename(knowledge = know, 
         acceptance_initial = accept, 
         acceptance_confirmation = confirm) %>% 
  # make a final acceptance variable
  mutate(acceptance = ifelse(is.na(acceptance_confirmation), acceptance_initial, acceptance_confirmation))
```

@@

# make data frame with average particpant data and wide format
d_wide <- left_join(d_wide, d_avg, 
               by = "id")
```

### 2024-04-17 exp_3/preregistration.Rmd
preregistration study 3

---
title: "Pre-registration Study 2 \n Who does not trust basic science?"
output:
  bookdown::pdf_document2: default
  bookdown::word_document2: default
  bookdown::html_document2:
      keep_md: yes
always_allow_html: true
bibliography: ../references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
```

@@

In a first study (https://osf.io/8utsj/) we tested whether participants would accept the scientific consensus on basic science facts. Most participants did, but not for all the questions. In this survey, we make two main adjustments to further investigate rejection of consensus: (i) we add a longer explanation and sources, instead of only stating the consensus. If it was the case that participants did not mistrust the science, but us the experimenters, this should increase acceptance; (ii) every time a participant rejects the scientific consensus, we ask them an open-ended question to explain why they did so. We also remove one science question that we had added ourselves (on where trees mainly draw there materials from to grow; answer: air), because the results were very different compared to other questions, suggesting the question might have been confusing.

@@

Next, participants answer a set of 10 basic science questions, all but one extracted from existing science knowledge questionnaires. After each question, participants will be presented with an answer reflecting the scientific consensus, along with an explanation which we wrote, partly based on explanations generated by ChatGPT, and three links to authoritative sources supporting the answer. Participants are asked to choose whether they accept the answer or not, before proceeding to the next question. Figure 1 displays the survey for an example science question. Table 1 shows all included questions, their scientifically consensual answer, the explanations, and their source.

@@

```{r stimulus-example, echo=FALSE, out.width= "100%", fig.align="center", fig.show="hold", fig.cap="(ref:stimulus-example)"}
knitr::include_graphics("materials/study2_question_example.png")
```

@@

```{r knowledge}
items <- read_csv("materials/knowledge_items_2.csv") %>% 
  mutate(id = 1:nrow(.)) %>% 
  select(id, Question, `Scientific consensus`, `Explanation`)

# Output the table
kbl(items, booktabs = T, longtable = TRUE,
    caption = "Science knowledge items", linesep = "\\addlinespace") %>%
  kable_paper(full_width = T) %>% 
  column_spec(1, width = "1em") %>%
  column_spec(2, width = "9em") %>%
  column_spec(3, width = "10em") %>%
  column_spec(4, width = "20em") 
```
## Conspiracy scales

@@

-   **RQ1: What is the average science knowledge score (1)?**

@@

In light of the previous study, we change RQ2 to:

-  **RQ2: When a participant’s answer does not match the consensus, how often do they change their mind and accept the consensus (2)?**

Similar to RQ1, we will report the average acceptance rate of the scientific consensus, when it was not already known. We will look at the pooled average and the distribution of participant averages.

@@

We will conduct a qualitative analysis and, if there are enough answers, attempt to categorize them.

In light of the results to the first study, we formulate the following hypotheses:

-  **H1a: Higher trust in science is associated with more science knowledge.**
-  **H1b: Higher trust in science is associated with more acceptance of the scientific consensus, given the consensus is not already known.**

We will regress trust in science on the average knowledge and acceptance (for those who initially disagreed) per participant. We will standardize all measures such that the regression coefficient yields a the correlation.

We will focus on the “How much do you trust science?” question, but conduct robustness checks on the two other trust questions.

We will also check how well the three trust items correlate.

-  **H2a: Higher conspiracy thinking is associated with more science knowledge.**
-  **H2b: Higher conspiracy thinking is associated with more acceptance of the scientific consensus, given the consensus is not already known.**

We will proceed just as for H1, but with conspiracy thinking as predictor, instead of trust.

To measure conspiracy thinking, we will use the average score of the Conspiracy Theory Inventory (BCTI) by Pennycook, Binnendyk, and Rand (2022). As robustness checks, we will run the same analysis separately for the two other conspiracy scales.

We will also check how well the three conspiracy scales correlate.

@@

We will recruit 200 participants from the US.

>>>>

---
title: "Pre-registration Study 3 \n Who does not trust basic science?"
output:
  bookdown::pdf_document2:
   extra_dependencies: ["float"]
  bookdown::word_document2: default
  bookdown::html_document2:
      keep_md: yes
always_allow_html: true
bibliography: ../references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.pos="H", echo = FALSE, message = FALSE)
```

@@

In two previous studies (https://osf.io/8utsj/) we tested whether participants would accept the scientific consensus on basic science facts. We first asked participants to answer basic science questions in a multiple choice format. After they had provided their answer, we presented participants with the scientific consensus on the question. Most participants did accept this consensus, even after having provided the wrong answers - but not for all the questions. In the second study, we provided participants with more extensive explanations and sources for the scientific consensus, as we anticipated that this might increase acceptance.  For cases in which participants still did not accept the consensus, we asked them to explain why. Both study 1 and 2 were run on convenience samples. Here, we will run study 2-–with some minor modifications-–on a sample of people holding anti-vaccination beliefs (see below). By contrast to study 2, after asking participants an open question about why they did not accept the consensus (in case they didn’t), we provide them with an explicit opportunity to change their answer. Based on the answers to the open questions we got in study 2, we also propose a categorization scheme of reasons why people rejected the consensus.

@@

Next, participants answer a set of 10 basic science questions, all but one extracted from existing science knowledge questionnaires. After each question, participants will be presented with an answer reflecting the scientific consensus, along with an explanation which we wrote, partly based on explanations generated by ChatGPT, and three links to authoritative sources supporting the answer. Participants are asked to choose whether they accept the answer or not, before proceeding to the next question. Figure \@ref(fig:stimulus-example) provides an example for one of the science questions. Table \@ref(tab:knowledge) shows all included questions, their scientifically consensual answer, the explanations, and their source.

@@

```{r stimulus-example, echo=FALSE, out.width= "50%", fig.align="center", fig.show="hold", fig.cap="(ref:stimulus-example)"}
knitr::include_graphics("materials/study2_question_example.png")
```

After answering all questions, if they had rejected the scientific consensus in some cases, we ask participants to explain why for each of these cases.  We also offer them the possibility to change their answer (Fig. \@ref(fig:explanation-example)).

(ref:explanation-example) Example of an explanation question and the opportunity to change the previous answer.

```{r explanation-example, echo=FALSE, out.width= "50%", fig.align="center", fig.show="hold", fig.cap="(ref:explanation-example)"}
knitr::include_graphics("materials/example_explanation_study3.png")
```

@@

```{r knowledge}
items <- read_csv("materials/knowledge_items_3.csv") %>% 
  mutate(id = 1:nrow(.)) %>% 
  select(id, Question, `Scientific consensus`, `Explanation`)

# Output the table
kbl(items, booktabs = T, longtable = TRUE,
    caption = "Science knowledge items", linesep = "\\addlinespace") %>%
  kable_paper(full_width = T) %>% 
  column_spec(1, width = "1em") %>%
  column_spec(2, width = "9em") %>%
  column_spec(3, width = "10em") %>%
  column_spec(4, width = "20em") 
```

## Conspiracy scales

@@

-   **RQ1: What is the average science knowledge score?**

@@

-  **RQ2: What is the average acceptance of the scientific consensus**

Similar to RQ1, we will report the average acceptance rate of the scientific consensus. We will look at the pooled average and the distribution of participant averages. Note that we will proceed as in study 1, i.e. we will not condition on initially false answers as we did in study 2.

@@

Based on the results of our last study, we propose 5 categories of justification for the rejection of the scientific consensus:

1. No justification provided/no clear justification (e.g. “I read the article.“)
2. Mistake, de facto acceptance of the consensus (e.g. "I made an error in my selection.")
3. Not convinced by the explanation provided (e.g. "I feel like the excerpt that was provided did not definitively state the answer.")

*Motivated rejection*:

4. Motivated rejection due to personal convictions (e.g. "I do not believe in the heliocentric model of an earth spinning around a centralized sun.")

5. Motivated rejection for religious reasons  (e.g. "I believe God make it they way it is and if He wanted it changed, He would have done it")

When answers fall in category 2, we expect people to change their answers and accept the scientific consensus when asked a second time. Those answers will be counted in the analysis as acceptance of the consensus.

# Hypotheses

In light of the results to the previous studies, we formulate the following hypotheses:

-  **H1a: Higher trust in science is associated with more science knowledge.**
-  **H1b: Higher trust in science is associated with more acceptance of the scientific consensus.**

We will regress trust in science on the average knowledge and acceptance per participant. We will focus on the “How much do you trust science?” question.
We will run a robustness check on the second Wellcome Global Monitor trust question, on the Pew question, and a combined measure of all three question items. We will also check how well the three trust items correlate.

-  **H2a: Higher conspiracy thinking is associated with less science knowledge.**
-  **H2b: Higher conspiracy thinking is associated with less acceptance of the scientific consensus.**

We will proceed just as for RQ3, but with conspiracy thinking as predictor, instead of trust. To measure conspiracy thinking, we will use the average score of the Conspiracy Theory Inventory (BCTI) by @pennycookOverconfidentlyConspiratorialConspiracy2022. As robustness checks, we will run the same analysis separately for the two other conspiracy scales (also averaging across all items). We will also check how well the three conspiracy scales correlate.

@@

We will recruit 200 participants from the US. To select a sample of participants distrusting vaccines, we combined three specific sample filters on  Prolific:

1. **COVID-19 Vaccine Opinions**: "Participants were asked the following question: Please describe your attitudes towards the COVID-19 (Coronavirus) vaccines:

  - For (I feel positively about the vaccines)
  - Against (I feel negatively about the vaccines)
  - Neutral (I don't have strong opinions either way)
  - Prefer not to say"

    We select only people who answered "Against".

2. **COVID-19 Vaccination**: "Participants were asked the following question: Have you received a coronavirus (COVID-19) vaccination?"

  - Yes (at least one dose)
  - No
  - Prefer not to answer"

    We select only people who answered "No".

3. **Vaccine Opinions 2**: "Participants were asked the following question: On a scale from 1-7, please rate to what extent you agree with the following statement: I believe that scheduled immunizations are safe for children.

 - 1 (TOTALLY DISAGREEE)
 - 2 (DISAGREE)
 - 3 (SOMEWHAT DISAGREE)
 - 4 (NEITHER AGREE NOR DISAGREE)
 - 5 (SOMEWHAT AGREE)
 - 6 (AGREE) 
 - 7 (TOTALLY AGREE)
 - Rather not say"

    We select only people who answered "1", "2", or "3".

The total sample available on Prolific with those three criteria combined is 667. If this pool of participants ist too small and we don’t reach the targeted sample size of 200 within a week of data collection, we’ll first remove the criterion “1. COVID-19 Vaccine Opinions”. If within the next three days we still don’t reach the targeted sample size we will also remove the criterion “2. COVID-19 Vaccination”.
